---
title: "Data Inlining"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data Inlining}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
# Skip if the duckdb R package is too old to support the ducklake v1.0 extension
duckdb_version_ok <- tryCatch({
  packageVersion("duckdb") >= "1.5.1"
}, error = function(e) FALSE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = duckdb_version_ok
)
# Use a unique temp directory for this vignette to avoid conflicts during R CMD check
vignette_temp_dir <- file.path(tempdir(), "data_inlining_vignette")
dir.create(vignette_temp_dir, showWarnings = FALSE, recursive = TRUE)
knitr::opts_knit$set(root.dir = vignette_temp_dir)
```

Data inlining is one of the flagship features of DuckLake v1.0.
When writing small changesets, DuckLake stores the data directly in the catalog
database instead of creating individual Parquet files. This eliminates the
"small files problem" that plagues traditional data lake formats during
streaming and frequent-update workloads.

Inlining is **enabled by default** with a row threshold of 10.
No configuration is required to benefit from it.

```{r setup, message = FALSE}
library(ducklake)
library(dplyr)
```

## How it works

Every insert, delete, or update that affects fewer rows than the inlining
threshold is stored in a lightweight table inside the catalog (DuckDB, SQLite
or PostgreSQL) instead of writing a Parquet file to the data path.
At query time DuckLake seamlessly combines inlined and Parquet data, so results
are always correct regardless of where data lives.

When you're ready, you can **flush** the inlined data to consolidated Parquet
files with a single call.

## Setup

```{r create-lake, message = FALSE}
install_ducklake()
attach_ducklake("sensor_lake", lake_path = vignette_temp_dir)
```

## Small writes are automatically inlined

Let's create a small sensor-readings table. Because the data has only 3 rows —
well below the default inlining threshold of 10 — DuckLake stores it directly
in the catalog instead of writing a Parquet file.

```{r create-table}
readings <- data.frame(
  sensor_id = 1:3,
  temperature = c(21.5, 22.1, 21.8),
  ts = as.POSIXct(c("2025-03-27 10:00:00", "2025-03-27 10:00:10", "2025-03-27 10:00:20"))
)

with_transaction(
  create_table(readings, "readings"),
  author = "Sensor Team",
  commit_message = "Initial sensor readings"
)
```

No Parquet files were created — all data is inlined in the catalog:

```{r verify-no-parquet}
conn <- get_ducklake_connection()
DBI::dbGetQuery(
  conn,
  sprintf("SELECT count(*) AS parquet_files FROM glob('%s/**/*.parquet');", vignette_temp_dir)
)
```

Yet queries return all rows seamlessly:

```{r query-inline}
get_ducklake_table("readings") |> collect()
```

## Incremental updates stay inlined

Small modifications — adding a derived column, correcting a value — also stay
in the catalog when the resulting changeset is below the threshold.

```{r small-update}
# Add a calibrated temperature column
with_transaction(
  get_ducklake_table("readings") |>
    mutate(temp_calibrated = temperature - 0.3) |>
    replace_table("readings"),
  author = "Sensor Team",
  commit_message = "Add calibrated temperature"
)

get_ducklake_table("readings") |> collect()
```

## Removing rows

Filtering out rows and replacing the table also works within the inlining
threshold:

```{r small-delete}
# Remove sensor 2's reading
with_transaction(
  get_ducklake_table("readings") |>
    filter(sensor_id != 2) |>
    replace_table("readings"),
  author = "Sensor Team",
  commit_message = "Remove faulty sensor 2 reading"
)

get_ducklake_table("readings") |> collect()
```

## Large writes bypass inlining automatically

When a write exceeds the threshold, DuckLake writes directly to Parquet — no
configuration needed:

```{r large-insert}
# 50 rows — well above the default threshold of 10
large_batch <- data.frame(
  sensor_id = 101:150,
  temperature = rnorm(50, mean = 22, sd = 1),
  ts = seq(as.POSIXct("2025-03-28 00:00:00"), by = "10 sec", length.out = 50),
  temp_calibrated = rnorm(50, mean = 21.7, sd = 1)
)

with_transaction(
  create_table(large_batch, "readings_bulk"),
  author = "Sensor Team",
  commit_message = "Bulk sensor upload"
)

# This table has a Parquet file
DBI::dbGetQuery(
  conn,
  sprintf("SELECT count(*) AS parquet_files FROM glob('%s/**/*.parquet');", vignette_temp_dir)
)
```

## Configuring the inlining threshold

### Global default

Change the threshold for all tables in the session:
```{r set-global}
# Increase the threshold for a streaming workload
set_inlining_row_limit(50)
get_inlining_row_limit()
```

### Per-connection at attach time

Set the threshold when attaching a DuckLake (not persisted):
```{r attach-limit, eval = FALSE}
attach_ducklake(
  "streaming_lake",
  lake_path = "/data/streaming",
  data_inlining_row_limit = 100
)
```

### Per-table persistent override

Set a table-level limit that is stored in the catalog and survives reconnects:
```{r table-limit, eval = FALSE}
set_inlining_row_limit(100, table_name = "readings")
```

### Disable inlining

```{r disable-inline}
set_inlining_row_limit(0)
# All writes now go directly to Parquet, even single rows
```

```{r restore-default, include = FALSE}
# Restore default for the rest of the vignette
set_inlining_row_limit(10)
```

## Flushing inlined data to Parquet

When inlined data accumulates, flush it to consolidated Parquet files:

```{r flush}
flush_result <- flush_inlined_data()
flush_result
```

Data remains correct after flushing:

```{r post-flush}
get_ducklake_table("readings") |> collect()
```

### Flushing a specific table

```{r flush-table, eval = FALSE}
flush_inlined_data(table_name = "readings")
```

## Checkpoint: one-stop maintenance

`checkpoint_ducklake()` runs all maintenance operations in sequence—flush,
compaction, snapshot expiration, and file cleanup:

```{r checkpoint}
checkpoint_ducklake()
```

Run checkpoints periodically (e.g., after a batch of streaming inserts) to
keep query performance optimal and consolidate inlined data.

## Time travel with inlined data

Inlined data fully supports DuckLake's time-travel capabilities. Each inlined
insert or delete creates a snapshot, just like a regular write:

```{r time-travel-setup}
# Check available snapshots
snapshots <- list_table_snapshots("readings")
snapshots
```

```{r time-travel-query}
# Query an earlier version
if (nrow(snapshots) > 0) {
  first_version <- snapshots$snapshot_id[1]
  get_ducklake_table_version("readings", first_version) |> collect()
}
```

## When to use inlining

| Workload | Recommendation |
|---|---|
| Streaming / IoT sensors | Increase threshold (e.g., 50–100) |
| Periodic large batch loads | Default (10) is fine |
| Single-row corrections | Default handles it automatically |
| Append-only bulk ETL | Consider disabling (`limit = 0`) |

## What inlining does _not_ change

Data inlining is a **storage optimisation**, not a change to DuckLake's data
integrity model. Whether data is inlined or written to Parquet:
- ACID transactions are fully enforced.
- Every modification creates a snapshot with author, timestamp, and commit
  message metadata.
- Time-travel queries work identically.
- The audit trail is complete and unaffected.

The only difference is _where_ the data physically resides until it is flushed:
in the catalog database (inlined) or in Parquet files on the data path.

### Regulated environments

In settings governed by ICH E6(R2), 21 CFR Part 11, or similar guidelines,
inlining is fully compatible with compliance requirements. Clinical trial
workflows typically involve batch loads of full SDTM or ADaM domains, which
exceed the default threshold and go directly to Parquet. Inlining primarily
benefits small targeted corrections (e.g., updating a single subject's flag),
keeping those lightweight without generating unnecessary files.

Before archival or regulatory submission, run `checkpoint_ducklake()` (or
`flush_inlined_data()`) to materialise all inlined data to Parquet. This
ensures that the data path contains a fully self-describing set of Parquet
files suitable for long-term storage and portability. See
`vignette("clinical-trial-datalake")` for a full regulated-workflow example.

## Summary

- **Inlining is on by default** (threshold = 10 rows). No setup required.
- Small inserts, deletes, and updates are stored in the catalog.
- Use `flush_inlined_data()` or `checkpoint_ducklake()` when ready to
  materialise to Parquet.
- Full time-travel support is preserved for inlined data.
- Adjust the threshold with `set_inlining_row_limit()` or at attach time via
  `data_inlining_row_limit`.

For more details, see the
[DuckLake data inlining documentation](https://ducklake.select/docs/stable/duckdb/advanced_features/data_inlining.html).

```{r cleanup, include = FALSE}
try(detach_ducklake("sensor_lake"), silent = TRUE)
```
