---
title: "Data Checks"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data Checks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
# Evaluate chunks only where the ducklake DuckDB extension is already
# installed. The probe never downloads anything, so building this vignette
# needs no network access. jsonlite writes the check results onto a commit.
ducklake_available <- ducklake::ducklake_extension_available() &&
  requireNamespace("jsonlite", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = ducklake_available
)
# Use a unique temp directory for this vignette to avoid conflicts during R CMD check
vignette_temp_dir <- file.path(tempdir(), "checks_vignette")
dir.create(vignette_temp_dir, showWarnings = FALSE, recursive = TRUE)
knitr::opts_knit$set(root.dir = vignette_temp_dir)
```

```{r setup, message=FALSE}
library(ducklake)
library(dplyr)

attach_ducklake("checks_lake", lake_path = vignette_temp_dir)

df_cars <- data.frame(model = rownames(mtcars), mtcars, row.names = NULL)

with_transaction(
  create_table(df_cars, "cars"),
  author = "Data Engineer",
  commit_message = "Add the Motor Trend car data"
)
```

DuckLake enforces one rule about the values in a table: a column can be
`NOT NULL`. It has no check constraints, no primary or unique keys, and no
foreign keys, and its catalog accepts no custom tags from SQL, so a rule
like "`cyl` is 4, 6, or 8" has no slot of its own. The catalog does hold
views. A view that returns the rows breaking a rule is a check: the rule's
logic sits in the catalog, it is versioned with the data it describes, and
any client of the lake runs it by reading the view.

This article writes checks that way and runs them with `run_checks()`. Then
it uses what a lake adds: a load that rolls back when a check fails, the
outcome of the checks recorded on a commit, and the rules and the data read
together as of an earlier snapshot. `run_checks()` is experimental while
the convention settles.

## The rule the lake enforces

`set_column_not_null()` sets the one constraint DuckLake has. From then on
the lake itself refuses a row without the value, whichever client writes
it. Plain SQL stands in for another client here:

```{r not_null}
with_transaction(
  set_column_not_null("cars", "mpg"),
  author = "Data Manager",
  commit_message = "Require mpg"
)

try(
  DBI::dbExecute(
    get_ducklake_connection(),
    "INSERT INTO cars (model, cyl) VALUES ('Unknown', 4)"
  )
)
```

## Rules as views

Every other rule is a view. The checks get a schema of their own (a schema
is the lake's namespace for tables and views), which keeps them apart from
the data and tells `run_checks()` where to look. Each one is a dplyr
pipeline that keeps the rows breaking the rule, stored with `create_view()`
and labelled with `set_table_comment()`:

```{r write_checks}
with_transaction({
  create_schema("checks")

  get_ducklake_table("main.cars") |>
    filter(!(cyl %in% c(4, 6, 8))) |>
    select(model, cyl) |>
    create_view("checks.cyl_known")
  set_table_comment("checks.cyl_known", "cyl is 4, 6, or 8")

  get_ducklake_table("main.cars") |>
    filter(!(hp > 0 & hp < 500)) |>
    select(model, hp) |>
    create_view("checks.hp_plausible")
  set_table_comment("checks.hp_plausible", "hp is between 0 and 500")
}, author = "Data Manager", commit_message = "Add the first checks on cars")
```

Four habits keep a check dependable:

- State the rule and negate it: `filter(!(rule))`. In SQL, `NOT` is not
  true for a row where the rule evaluates to `NA`, so a missing `cyl` passes
  `cyl_known`. When a missing value matters, give it a check of its own, or
  make the column `NOT NULL`.
- Select the columns a reviewer needs. The view is the listing of what
  failed, so its columns are the ones someone will act on.
- Read the table by its schema-qualified name, `"main.cars"`. DuckDB
  resolves an unqualified name from the view's schema and the session's
  current database, so `"cars"` can fail to bind for a client where
  another database is current.
- Keep only checks in the schema. `run_checks()` counts the rows of every
  view it finds there, and a view that summarizes (a row of totals) would
  report a failure every time.

A rule that spans tables is a join. An `anti_join()` of visits against
subjects, for example, returns the visits with no subject.

## Running the checks

`run_checks()` counts the rows each check returns:

```{r run_checks}
run_checks()
```

The label comes from the view's comment and `n_fail` is the number of rows
breaking the rule. The car data passes both.

## Gating a load

A new batch arrives, and one of its rows has a typing error:

```{r batch}
df_batch <- data.frame(
  model = c("Saab 99", "Audi 100 LS"),
  mpg = c(24.5, 23.0),
  cyl = c(4, 40),
  hp = c(87, 91)
)
```

Inside a transaction the checks see the rows that are waiting to be
committed. So a load can run the checks on itself and stop, and
`with_transaction()` rolls everything back when it does:

```{r gate}
try(
  with_transaction({
    rows_insert(get_ducklake_table("cars"), df_batch, by = "model")

    df_checks <- run_checks()
    if (any(df_checks$n_fail > 0)) {
      stop("failed ", toString(df_checks$check[df_checks$n_fail > 0]))
    }
  }, author = "Data Manager", commit_message = "Load the September batch")
)
```

Nothing of the batch reached the lake. The table has its 32 rows and the
history has no new snapshot:

```{r gate_result}
get_ducklake_table("cars") |> count() |> collect()

list_table_snapshots("cars") |>
  select(snapshot_id, author, commit_message)
```

This is how a rule becomes a constraint in a lake. DuckLake would accept
the row; a transaction that checks itself never commits it. The gate covers
the writers that use it, and a client that writes without running the
checks is not stopped.

## Recording the outcome on the commit

Rejecting the batch is not always right. Raw data often has to land as it
arrived, with the problems flagged for follow-up. A manual transaction can
run the checks after the write and put the outcome on the commit, in
`commit_extra_info`:

```{r soft_load}
begin_transaction()
rows_insert(get_ducklake_table("cars"), df_batch, by = "model")

df_failed <- run_checks() |>
  filter(n_fail > 0) |>
  select(check, n_fail)

commit_transaction(
  author = "Data Manager",
  commit_message = "Load the September batch",
  commit_extra_info = as.character(
    jsonlite::toJSON(list(checks_failed = df_failed))
  )
)

v_loaded <- max(list_table_snapshots("cars")$snapshot_id)
```

The snapshot now says what state the data was in when it was committed:

```{r soft_result}
list_table_snapshots("cars") |>
  select(snapshot_id, commit_message, commit_extra_info) |>
  tail(1)
```

## Which rows failed

The check is a view, so the failing rows are one read away, with the
columns the check selected:

```{r listing}
run_checks()

get_ducklake_table("checks.cyl_known") |> collect()
```

The correction is an ordinary update, and its commit message can name the
check it answers. Afterwards the lake's history holds the load, the
finding, and the fix:

```{r fix}
with_transaction(
  rows_update(
    get_ducklake_table("cars"),
    data.frame(model = "Audi 100 LS", cyl = 4),
    by = "model"
  ),
  author = "Data Manager",
  commit_message = "cyl_known: Audi 100 LS cyl corrected from 40 to 4"
)

run_checks()
```

## Rules have history

A check is a catalog object, so changing a rule is a snapshot with an author
and a message, like a change to the data. Replacing a view makes a new
catalog entry, which starts without a comment. Set the label again in the
same transaction:

```{r revise_check}
with_transaction({
  get_ducklake_table("main.cars") |>
    filter(!(hp >= 40 & hp <= 400)) |>
    select(model, hp) |>
    create_view("checks.hp_plausible")
  set_table_comment("checks.hp_plausible", "hp is between 40 and 400")
}, author = "Data Manager", commit_message = "Narrow the plausible hp range")

list_table_snapshots("checks.hp_plausible") |>
  select(snapshot_id, author, commit_message)
```

## Rules and data as of a snapshot

A lake attached at a snapshot shows its views as they were then, along with
its tables. Re-attaching at the snapshot of the September load runs the
rules of that day against the data of that day. The uncorrected row fails
again, and the `hp` rule reads as it did then:

```{r pinned, message=FALSE}
detach_ducklake("checks_lake")
attach_ducklake(
  "checks_lake",
  lake_path = vignette_temp_dir,
  snapshot_version = v_loaded
)

run_checks()
```

That answers a question an audit asks: which rules were in force when this
data was committed, and did the data pass them?

## Where other tools fit

`run_checks()` counts and nothing more. A report is a job for the tools
built for one: the rows collected from a check view are a data frame, ready
for gt, or for a validation report from affirm or pointblank. pointblank
also interrogates lazy tables, so one of its agents can run on
`get_ducklake_table("cars")` inside DuckDB. For a dictionary of a lake's
tables and columns that travels as a file,
[data-dict](https://data-dict.tidyverse.org) defines a YAML format.
Comments and labels, the descriptive half of a dictionary, are covered in
`vignette("views-comments-labels")`.

```{r cleanup}
detach_ducklake("checks_lake")
```
