---
title: "Loading Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Loading Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
# Evaluate chunks only where the ducklake DuckDB extension is already
# installed. The probe never downloads anything, so building this vignette
# needs no network access.
ducklake_available <- ducklake::ducklake_extension_available()
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = ducklake_available
)
# Use a unique temp directory for this vignette to avoid conflicts during R CMD check
vignette_temp_dir <- file.path(tempdir(), "loading_data_vignette")
dir.create(vignette_temp_dir, showWarnings = FALSE, recursive = TRUE)
knitr::opts_knit$set(root.dir = vignette_temp_dir)
```

```{r setup, message=FALSE}
library(ducklake)
library(dplyr)

attach_ducklake("loading_lake", lake_path = vignette_temp_dir)
```

`create_table()` is the one entry point for data that is not in the lake
yet. It accepts a data frame, a path to a file, a URL, or a lazy table, and
in every case the result is a new table and a new snapshot. The recipes
below start with the everyday cases and end with the migration paths:
registering Parquet files that already exist, and copying whole databases
in from DuckDB or Iceberg.

Each load here is wrapped in `with_transaction()` so that the snapshot
carries an author and a message. [Getting Started](ducklake.html) explains
why.

## From a data frame

```{r data_frame}
with_transaction(
  create_table(mtcars, "cars"),
  author = "Data Engineer",
  commit_message = "Add the Motor Trend car data"
)
```

Factor columns become character columns on the way in, because DuckLake
has no `ENUM` type, and `create_table()` says so when it happens:

```{r factors}
with_transaction(
  create_table(iris, "flowers"),
  author = "Data Engineer",
  commit_message = "Add the iris measurements"
)
```

Variable labels (the `label` attributes that haven and labelled use) are
stored as column comments and come back on `collect()`.
[Views, Comments, and Labels](views-comments-labels.html) shows the round
trip.

## From a file

A file path goes straight to DuckDB's readers, so CSV, Parquet, and JSON
files all load with the same call:

```{r csv}
csv_path <- file.path(vignette_temp_dir, "sample_data.csv")
write.csv(head(iris, 20), csv_path, row.names = FALSE)

with_transaction(
  create_table(csv_path, "iris_sample"),
  author = "Data Engineer",
  commit_message = "Load the iris sample from CSV"
)

get_ducklake_table("iris_sample") |> head(3)
```

## From a URL

An `http://` or `https://` address works the same way. ducklake loads
DuckDB's `httpfs` extension for it:

```{r url, eval=FALSE}
with_transaction(
  create_table("https://example.com/data.csv", "remote_data"),
  author = "Data Engineer",
  commit_message = "Load the remote dataset"
)
```

## From a dplyr pipeline

A pipeline on an R data frame is collected and written like any other data
frame:

```{r pipeline}
with_transaction(
  mtcars |>
    filter(mpg > 20) |>
    create_table("efficient_cars"),
  author = "Data Analyst",
  commit_message = "Add the cars above 20 mpg"
)
```

## Derive a table inside the database

When the pipeline starts from a lake table, `create_table()` runs it as
`CREATE TABLE ... AS` inside DuckDB. The rows never pass through R, and
column labels follow the columns into the new table:

```{r derive}
with_transaction(
  get_ducklake_table("cars") |>
    filter(cyl == 4) |>
    select(mpg, cyl, hp, wt) |>
    create_table("small_cars"),
  author = "Data Analyst",
  commit_message = "Four-cylinder subset"
)
```

This is the shape of a medallion pipeline: each layer is a query on the
layer before it, written straight into the lake.

## Put tables in schemas

Schemas group tables inside the lake. They suit medallion layers
(`bronze`, `silver`, `gold`) or one area per study, and every function that
takes a table name accepts `"schema.table"`:

```{r schemas}
with_transaction({
  create_schema("staging")
  create_table(mtcars, "staging.cars_raw")
},
  author = "Data Engineer",
  commit_message = "Stage the raw car data"
)

get_ducklake_table("staging.cars_raw") |>
  count(cyl)

list_ducklake_tables()
```

## Register Parquet files in place

If your data is already in Parquet, `add_data_files()` records the files in
the lake where they are: no copy, no rewrite, and nothing collected into R.
A vector of files is registered atomically in one snapshot, which makes
this the fast migration path for a folder of Parquet extracts. The target
table can already exist with a compatible schema, or `create = TRUE`
creates it from the Parquet schema. The lake takes ownership of the files,
and later compaction may rewrite or delete them.

```{r add_data_files, eval=FALSE}
add_data_files(
  "readings",
  c("extracts/jan.parquet", "extracts/feb.parquet"),
  create = TRUE
)

# See which files back a table
list_ducklake_files("readings")
```

## Migrate a DuckDB database

A DuckDB database file moves into the lake with one statement: attach it
read-only and copy every schema and table across. The copy lands as a
single snapshot.

```{r migrate_duckdb}
legacy_db <- file.path(vignette_temp_dir, "legacy.duckdb")
legacy <- DBI::dbConnect(duckdb::duckdb(dbdir = legacy_db))
DBI::dbWriteTable(
  legacy, "sites",
  data.frame(site_id = 1:3, country = c("US", "DE", "JP"))
)
DBI::dbDisconnect(legacy, shutdown = TRUE)

conn <- get_ducklake_connection()
DBI::dbExecute(conn, sprintf("ATTACH '%s' AS legacy (READ_ONLY);", legacy_db))
DBI::dbExecute(conn, "COPY FROM DATABASE legacy TO loading_lake;")
DBI::dbExecute(conn, "DETACH legacy;")

get_ducklake_table("sites") |> collect()
```

Types DuckLake does not support (`ENUM`, `UNION`, `VARINT`, fixed-size
arrays) need converting first; the DuckLake documentation has a migration
script for those cases:
<https://ducklake.select/docs/stable/duckdb/migrations/duckdb_to_ducklake>.

## Exchange data with Iceberg

With DuckDB's iceberg extension attached to an Iceberg REST catalog,
`COPY FROM DATABASE` moves tables in either direction. Copying into Iceberg
needs the schemas to exist there first, and the copy adds tables rather
than replacing them.

```{r iceberg, eval=FALSE}
DBI::dbExecute(conn, "INSTALL iceberg; LOAD iceberg;")
DBI::dbExecute(conn, "
  ATTACH '' AS iceberg_catalog (
    TYPE iceberg,
    CLIENT_ID 'admin',
    CLIENT_SECRET 'password',
    ENDPOINT 'http://iceberg.example.org:8181'
  );
")

# Lake to Iceberg
DBI::dbExecute(conn, "COPY FROM DATABASE loading_lake TO iceberg_catalog;")

# Iceberg to lake: the data itself, or only the metadata, so that Iceberg
# tables read as lake tables where they are
DBI::dbExecute(conn, "COPY FROM DATABASE iceberg_catalog TO loading_lake;")
DBI::dbExecute(conn, "CALL iceberg_to_ducklake('iceberg_catalog', 'loading_lake');")
```

```{r cleanup}
detach_ducklake("loading_lake")
```
