---
title: "Modifying Tables with Version Control"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Modifying Tables with Version Control}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
# Evaluate chunks only where the ducklake DuckDB extension is usable: needs
# duckdb >= 1.5.1 and, when not already cached, network access to install the
# extension (not available on CRAN).
ducklake_available <- tryCatch({
  packageVersion("duckdb") >= "1.5.1" && {
    con <- DBI::dbConnect(duckdb::duckdb())
    ok <- tryCatch({
      DBI::dbExecute(con, "INSTALL ducklake;")
      DBI::dbExecute(con, "LOAD ducklake;")
      TRUE
    }, error = function(e) FALSE)
    DBI::dbDisconnect(con, shutdown = TRUE)
    ok
  }
}, error = function(e) FALSE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = ducklake_available
)
# Use a unique temp directory for this vignette to avoid conflicts during R CMD check
vignette_temp_dir <- file.path(tempdir(), "modifying_tables_vignette")
dir.create(vignette_temp_dir, showWarnings = FALSE, recursive = TRUE)
knitr::opts_knit$set(root.dir = vignette_temp_dir)
```

This vignette demonstrates how to modify tables in a DuckLake while maintaining complete version control and audit trails. 
This is essential for reproducible workflows.

```{r setup, message = FALSE}
library(ducklake)
library(dplyr)

# Setup for examples
install_ducklake()
attach_ducklake("modifying_tables_lake", lake_path = vignette_temp_dir)

# Load a sample dataset
with_transaction(
  create_table(mtcars, "cars"),
  author = "Data Engineer",
  commit_message = "Initial car data load"
)
```

A note before we start: when you load dplyr you may see a message that it
masks ducklake's `rows_insert()`, `rows_update()`, and `rows_delete()`. That
is harmless. Tables returned by `get_ducklake_table()` carry a class that
dispatches to ducklake's DuckLake-aware methods regardless of the order in
which the packages were loaded.

## Choosing a Modification Approach

Good news first: **every committed change to a DuckLake table creates a
snapshot**. Whether you use `rows_insert()`, `replace_table()`, or raw SQL,
DuckLake records what changed and you can time-travel back to any earlier
state. (Earlier versions of this vignette said the `rows_*` functions skip
versioning -- that is not true in DuckLake v1.0.) The choice between the
styles is about *what kind of change* you are making, not about whether it is
audited.

One distinction matters before anything else: dplyr joins **read** --
`left_join()` and friends combine tables into a new result and never touch
the stored data. Everything else in this table **writes**.

| You want to... | Reach for |
|---|---|
| Look up or combine data for analysis | dplyr joins (`left_join()`, ...) |
| Append, correct, or remove specific rows | `rows_insert()`, `rows_update()`, `rows_delete()` |
| Update rows that exist, insert the ones that don't | `rows_upsert()` |
| Conditional updates, or deletes driven by a staging table | `merge_into()` |
| Bulk transformations that touch most rows | `replace_table()` |

### For incremental changes: the `rows_*` functions

Use `rows_insert()`, `rows_update()`, and `rows_delete()` when you are
appending records, correcting specific values, or removing specific rows:

```r
# Each of these is one SQL statement and one new snapshot
rows_insert(get_ducklake_table("my_table"), new_data, by = "id")
rows_update(get_ducklake_table("my_table"), corrections, by = "id")
rows_delete(get_ducklake_table("my_table"), obsolete_ids, by = "id")
```

**Why they shine for incremental work:**

- **Efficient** - the change runs inside DuckDB as a single statement; the
  rest of the table is never read into R or rewritten
- **Streaming-friendly** - small changes benefit from DuckLake's
  [data inlining](data-inlining.html), landing in the catalog instead of
  spawning tiny Parquet files
- **Still versioned** - each call produces a snapshot you can time-travel to

### For update-or-insert: `rows_upsert()`

When a batch mixes corrections to existing rows with rows you have never
seen, `rows_upsert()` handles both in one atomic statement: rows whose key
matches are updated, the rest are inserted.

```r
# One statement, one snapshot: id 2 is updated, id 7 is inserted
rows_upsert(get_ducklake_table("my_table"), mixed_batch, by = "id")
```

DuckLake tables have no primary keys, so under the hood this is SQL
`MERGE INTO` matching on your `by` columns, not the `ON CONFLICT` upsert you
may know from other databases. The practical upshot is the same, with one
wrinkle: when the upsert data covers only some of the table's
columns, *inserted* rows get the column default (usually `NULL`) in the
columns you didn't supply, while *updated* rows keep their existing values
there.

### For staging syncs and conditional merges: `merge_into()`

`rows_upsert()` covers update-or-insert. `merge_into()` exposes the rest of
the SQL MERGE statement for the cases beyond it:

```r
# Update only when the source is newer
merge_into(
  get_ducklake_table("my_table"), fresh_data, by = "id",
  matched_condition = "source.updated_at > target.updated_at"
)

# Synchronize to a staging table: upsert, plus delete rows
# that no longer exist in the source
merge_into("my_table", staging, by = "id", delete_missing = TRUE)
```

Despite the name, `merge_into()` is unrelated to `base::merge()` and
dplyr's joins, which combine tables into a new result. `merge_into()`
changes the target table in place.

You could get a similar end state by joining the staging table to the
current table and calling `replace_table()` on the result. Resist that
instinct: it rewrites every row of the table, and the change feed then
records a wholesale replacement instead of the handful of inserts, updates,
and deletes that actually happened. `merge_into()` touches only the affected
rows, so `get_table_changes()` afterward tells the true story.

### For structural or bulk changes: `replace_table()`

Use `replace_table()` when the *shape* of the table changes -- adding or
removing columns -- or when a transformation touches most rows anyway:

```r
with_transaction(
  get_ducklake_table("my_table") |>
    filter(status == "active") |>
    mutate(processed = TRUE) |>
    replace_table("my_table"),
  author = "Your Name",
  commit_message = "Mark active records as processed"
)
```

`replace_table()` collects the transformed data into R and rewrites the
table, which is exactly right for schema changes but wasteful for touching
three rows in a million-row table.

### Group related changes with `with_transaction()`

Whichever style you use, wrap *related* modifications in
`with_transaction()`. All changes inside the transaction become **one**
snapshot, and you can attach an author and commit message for the audit
trail -- valuable in any setting and essential for GxP/21 CFR Part 11 work:

```r
with_transaction({
  rows_insert(get_ducklake_table("my_table"), march_batch, by = "id")
  rows_delete(get_ducklake_table("my_table"), recalled_units, by = "id")
},
  author = "Data Team",
  commit_message = "March intake; remove recalled units"
)
```

## Examples

### Incremental changes with the `rows_*` functions

Let's see the row-level functions in action on a small fleet table with a
proper key column:

```{r rows-setup}
fleet <- data.frame(
  car_id = 1:3,
  model = c("Corolla", "Civic", "Model 3"),
  mileage = c(42000, 38500, 12000)
)

with_transaction(
  create_table(fleet, "fleet"),
  author = "Fleet Manager",
  commit_message = "Initial fleet inventory"
)
```

**Insert** new records by key. The new rows are appended in a single SQL
statement -- the existing rows are never read into R:

```{r rows-insert}
new_cars <- data.frame(
  car_id = 4:5,
  model = c("Leaf", "Ioniq 5"),
  mileage = c(500, 120)
)

rows_insert(get_ducklake_table("fleet"), new_cars, by = "car_id")

get_ducklake_table("fleet") |> collect()
```

**Update** specific values by key. Only the matched rows change:

```{r rows-update}
correction <- data.frame(car_id = 2, mileage = 39000)

rows_update(get_ducklake_table("fleet"), correction, by = "car_id")

get_ducklake_table("fleet") |> filter(car_id == 2) |> collect()
```

**Delete** rows by key:

```{r rows-delete}
sold <- data.frame(car_id = 1)

rows_delete(get_ducklake_table("fleet"), sold, by = "car_id")

get_ducklake_table("fleet") |> collect()
```

Each call above created its own snapshot. To record an author and commit
message -- or to make several row operations land as **one** snapshot -- wrap
them in `with_transaction()`:

```{r rows-transaction}
april_arrivals <- data.frame(car_id = 6, model = "ID.4", mileage = 60)
recalled <- data.frame(car_id = 4)

with_transaction({
  rows_insert(get_ducklake_table("fleet"), april_arrivals, by = "car_id")
  rows_delete(get_ducklake_table("fleet"), recalled, by = "car_id")
},
  author = "Fleet Manager",
  commit_message = "April intake; remove recalled Leaf"
)

# The full history: every change is versioned, wrapped or not
list_table_snapshots("fleet")
```

**Upsert** a batch that mixes corrections and new arrivals. The Model 3's
mileage is updated and the Kona is inserted, in one statement:

```{r rows-upsert}
service_batch <- data.frame(
  car_id = c(3, 7),
  model = c("Model 3", "Kona"),
  mileage = c(15200, 8000)
)

rows_upsert(get_ducklake_table("fleet"), service_batch, by = "car_id")

get_ducklake_table("fleet") |> arrange(car_id) |> collect()
```

**Synchronize** to an authoritative source with `merge_into()`. Suppose the
quarterly registry export is the truth: matching cars take its values, cars
it doesn't list are gone, and cars we haven't seen are added. One call, one
snapshot:

```{r merge-into}
registry <- data.frame(
  car_id = c(3, 5, 8),
  model = c("Model 3", "Ioniq 5", "e-Golf"),
  mileage = c(15400, 900, 21000)
)

with_transaction(
  merge_into("fleet", registry, by = "car_id", delete_missing = TRUE),
  author = "Fleet Manager",
  commit_message = "Quarterly registry sync"
)

get_ducklake_table("fleet") |> arrange(car_id) |> collect()
```

Because the sync ran as targeted row changes rather than a table rewrite,
the change feed records exactly what happened:

```{r merge-into-changes}
latest <- max(list_table_snapshots("fleet")$snapshot_id)
get_table_changes("fleet", latest, latest) |>
  select(change_type, car_id, model, mileage) |>
  collect()
```

### Updating specific rows with `replace_table()`

```{r update-rows}
# Update mpg values for specific cars (4-cylinder cars get a 5% efficiency boost)
with_transaction(
  get_ducklake_table("cars") |>
    mutate(
      mpg = if_else(cyl == 4, mpg * 1.05, mpg)
    ) |>
    replace_table("cars"),
  author = "Data Engineer",
  commit_message = "Update MPG for 4-cylinder vehicles"
)

# Check version history - should show the new snapshot
list_table_snapshots("cars")
```

### Adding derived columns

```{r add-columns}
# Add new derived columns to existing table
with_transaction(
  get_ducklake_table("cars") |>
    mutate(
      hp_per_cyl = hp / cyl,
      # Add a new flag column
      high_performance = if_else(hp > 200, "Y", "N")
    ) |>
    replace_table("cars"),
  author = "Data Engineer",
  commit_message = "Add HP per cylinder and performance flag"
)

# Verify new columns exist
get_ducklake_table("cars") |>
  filter(hp > 200) |>
  select(hp, cyl, hp_per_cyl, high_performance)
```

### Filtering rows with `replace_table()`

```{r filter}
# Keep only specific rows - creates a versioned snapshot
with_transaction(
  get_ducklake_table("cars") |>
    filter(cyl == 8) |>
    replace_table("cars"),
  author = "Data Engineer",
  commit_message = "Filter to V8 engines only"
)

# Show the filtered table
get_ducklake_table("cars")

# View version history - old versions still accessible via time travel
list_table_snapshots("cars")
```

### Time Travel: Accessing Previous Versions

```{r time-travel}
# Get the current version
current <- get_ducklake_table("cars") |> collect()

# List all snapshots to see available versions
snapshots <- list_table_snapshots("cars")
snapshots

# Access a specific previous version by snapshot_id
original_version <- get_ducklake_table_version(
  "cars", 
  snapshots$snapshot_id[1]
) |> collect()

# Compare: how many rows changed?
nrow(current)
nrow(original_version)
```

```{r cleanup, include=FALSE}
detach_ducklake("modifying_tables_lake")
```
