---
title: "Modifying Tables with Version Control"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Modifying Tables with Version Control}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
# Skip if the duckdb R package is too old to support the ducklake v1.0 extension
duckdb_version_ok <- tryCatch({
  packageVersion("duckdb") >= "1.5.1"
}, error = function(e) FALSE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = duckdb_version_ok
)
# Use a unique temp directory for this vignette to avoid conflicts during R CMD check
vignette_temp_dir <- file.path(tempdir(), "modifying_tables_vignette")
dir.create(vignette_temp_dir, showWarnings = FALSE, recursive = TRUE)
knitr::opts_knit$set(root.dir = vignette_temp_dir)
```

This vignette demonstrates how to modify tables in a DuckLake while maintaining complete version control and audit trails. 
This is essential for reproducible workflows.

```{r setup}
library(ducklake)
library(dplyr)

# Setup for examples
install_ducklake()
attach_ducklake("modifying_tables_lake", lake_path = vignette_temp_dir)

# Load a sample dataset
with_transaction(
  create_table(mtcars, "cars"),
  author = "Data Engineer",
  commit_message = "Initial car data load"
)
```

## Choosing a Modification Approach

Good news first: **every committed change to a DuckLake table creates a
snapshot**. Whether you use `rows_insert()`, `replace_table()`, or raw SQL,
DuckLake records what changed and you can time-travel back to any earlier
state. (Earlier versions of this vignette said the `rows_*` functions skip
versioning -- that is not true in DuckLake v1.0.) The choice between the two
styles is about *what kind of change* you are making, not about whether it is
audited.

### For incremental changes: the `rows_*` functions

Use `rows_insert()`, `rows_update()`, and `rows_delete()` when you are
appending records, correcting specific values, or removing specific rows:

```r
# Each of these is one SQL statement and one new snapshot
rows_insert(get_ducklake_table("my_table"), new_data, by = "id")
rows_update(get_ducklake_table("my_table"), corrections, by = "id")
rows_delete(get_ducklake_table("my_table"), obsolete_ids, by = "id")
```

**Why they shine for incremental work:**

- **Efficient** - the change runs inside DuckDB as a single statement; the
  rest of the table is never read into R or rewritten
- **Streaming-friendly** - small changes benefit from DuckLake's
  [data inlining](data-inlining.html), landing in the catalog instead of
  spawning tiny Parquet files
- **Still versioned** - each call produces a snapshot you can time-travel to

### For structural or bulk changes: `replace_table()`

Use `replace_table()` when the *shape* of the table changes -- adding or
removing columns -- or when a transformation touches most rows anyway:

```r
with_transaction(
  get_ducklake_table("my_table") |>
    filter(status == "active") |>
    mutate(processed = TRUE) |>
    replace_table("my_table"),
  author = "Your Name",
  commit_message = "Mark active records as processed"
)
```

`replace_table()` collects the transformed data into R and rewrites the
table, which is exactly right for schema changes but wasteful for touching
three rows in a million-row table.

### Group related changes with `with_transaction()`

Whichever style you use, wrap *related* modifications in
`with_transaction()`. All changes inside the transaction become **one**
snapshot, and you can attach an author and commit message for the audit
trail -- valuable in any setting and essential for GxP/21 CFR Part 11 work:

```r
with_transaction({
  rows_insert(get_ducklake_table("my_table"), march_batch, by = "id")
  rows_delete(get_ducklake_table("my_table"), recalled_units, by = "id")
},
  author = "Data Team",
  commit_message = "March intake; remove recalled units"
)
```

## Examples

### Updating specific rows with `replace_table()`

```{r update-rows}
# Update mpg values for specific cars (4-cylinder cars get a 5% efficiency boost)
with_transaction(
  get_ducklake_table("cars") |>
    mutate(
      mpg = if_else(cyl == 4, mpg * 1.05, mpg)
    ) |>
    replace_table("cars"),
  author = "Data Engineer",
  commit_message = "Update MPG for 4-cylinder vehicles"
)

# Check version history - should show the new snapshot
list_table_snapshots("cars")
```

### Adding derived columns

```{r add-columns}
# Add new derived columns to existing table
with_transaction(
  get_ducklake_table("cars") |>
    mutate(
      hp_per_cyl = hp / cyl,
      # Add a new flag column
      high_performance = if_else(hp > 200, "Y", "N")
    ) |>
    replace_table("cars"),
  author = "Data Engineer",
  commit_message = "Add HP per cylinder and performance flag"
)

# Verify new columns exist
get_ducklake_table("cars") |>
  filter(hp > 200) |>
  select(hp, cyl, hp_per_cyl, high_performance)
```

### Filtering rows with `replace_table()`

```{r filter}
# Keep only specific rows - creates a versioned snapshot
with_transaction(
  get_ducklake_table("cars") |>
    filter(cyl == 8) |>
    replace_table("cars"),
  author = "Data Engineer",
  commit_message = "Filter to V8 engines only"
)

# Show the filtered table
get_ducklake_table("cars")

# View version history - old versions still accessible via time travel
list_table_snapshots("cars")
```

### Time Travel: Accessing Previous Versions

```{r time-travel}
# Get the current version
current <- get_ducklake_table("cars") |> collect()

# List all snapshots to see available versions
snapshots <- list_table_snapshots("cars")
snapshots

# Access a specific previous version by snapshot_id
original_version <- get_ducklake_table_version(
  "cars", 
  snapshots$snapshot_id[1]
) |> collect()

# Compare: how many rows changed?
nrow(current)
nrow(original_version)
```

```{r cleanup, include=FALSE}
detach_ducklake("my_ducklake")
```
