---
title: "Quack Remote Access"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quack Remote Access}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  # Quack needs a running server and DuckDB 1.5.3 or newer, neither of which is
  # available when this vignette is built. The examples are shown with their
  # expected output rather than executed.
  eval = FALSE
)
```

```{r setup}
library(ducklake)
library(dplyr)
```

## Introduction

Most of the time a DuckLake lives in one place and one person works with it. But
data analysis is usually a team activity: a few statisticians share the same
study data, an analytics group works from one set of cleaned tables, several
people review the same numbers before a deliverable goes out. The usual answer
is to copy files around (a CSV here, an XPT there) and hope everyone is looking
at the same version.

[Quack](https://duckdb.org/quack/) offers a different answer. It is a protocol,
shipped as a core DuckDB extension in DuckDB 1.5.3, that lets one DuckDB instance
act as a server while other DuckDB instances connect to it as clients. A
DuckLake held by the server can then be read and updated by other R sessions over
the network, with the same snapshots and time travel you get locally. Several
people can write at the same time without locking each other out.

This vignette shows the two halves of that picture: connecting to a lake someone
else is serving, and serving a lake of your own. The audience here is the
everyday R user, so the examples stay close to ordinary analysis work and do not
assume any server administration experience.

The code below is shown with its expected output but is not run when the vignette
is built, because Quack needs a live server and a recent DuckDB. To follow along
you need the `duckdb` R package at version 1.5.3 or newer:

```{r version-check}
packageVersion("duckdb")
#> [1] '1.5.3'
```

## Installing the extension

Quack is a core extension, so DuckDB loads it automatically the first time it is
used. On a machine that has no internet access at query time you can install it
ahead of time:

```{r install}
install_quack()
#> Installed quack extension.
#> Loaded quack extension.
```

## Connecting to a shared lake

This is the common case. A colleague, or a shared workstation, runs a Quack
server that holds the team's DuckLake. You connect to it from your own R session
and pull the data you need.

`quack_query()` sends a single SQL query to the server and returns the result as
a data frame. The lake is named by its catalog on the server, which is the name
it was attached under (here, `trial`):

```{r read}
adsl <- quack_query(
  "quack:trial-server.example.org",
  "SELECT * FROM trial.adsl",
  token = "super_secret"
)

adsl
#>   USUBJID SAFFL AGE
#> 1  01-001     Y  45
#> 2  01-002     N  52
```

The address starts with `quack:`. A bare host name works too and the scheme is
added for you. The default port is 9494; include a different one as
`quack:host:port`. The `token` is whatever password the server was started with.

Once a query comes back you have an ordinary data frame, so the rest of your
analysis is the dplyr and base R you already use:

```{r analyze}
adsl |>
  group_by(SAFFL) |>
  summarise(n = n(), mean_age = mean(AGE))
#> # A tibble: 2 x 3
#>   SAFFL     n mean_age
#>   <chr> <int>    <dbl>
#> 1 N         1       52
#> 2 Y         1       45
```

For large tables it is faster to let the server do the filtering and summarising
and send back only what you need, rather than pulling every row across the
network:

```{r read-filtered}
quack_query(
  "quack:trial-server.example.org",
  "SELECT USUBJID, AGE FROM trial.adsl WHERE SAFFL = 'Y' AND AGE >= 18",
  token = "super_secret"
)
#>   USUBJID AGE
#> 1  01-001  45
```

Because Quack supports concurrent writers, you can also add data to the shared
lake. The query runs on the server, so it refers to the lake by its catalog
name and returns the number of rows affected:

```{r write}
quack_query(
  "quack:trial-server.example.org",
  "INSERT INTO trial.adsl VALUES ('01-099', 'Y', 70)",
  token = "super_secret"
)
#>   Count
#> 1     1
```

### Connecting to a plain database

If the server is sharing a regular DuckDB database rather than a DuckLake, you
can attach it as a catalog and work with it through the familiar
`get_ducklake_table()` and dplyr path. `attach_quack()` gives the remote a local
name; its tables are then reached as `name.table`:

```{r attach}
attach_quack("warehouse", "quack:data.example.org", token = "super_secret")

get_ducklake_table("warehouse.sales") |>
  filter(region == "EMEA") |>
  collect()

detach_quack("warehouse")
```

A served DuckLake is a separate catalog on the server, not part of its default
database, so its tables are not exposed this way. Reach a served DuckLake with
`quack_query()` and the lake's catalog name, as in the previous section.

## Serving a lake of your own

The other half of the picture is publishing a lake so others can connect to it.
Say you have built a small medallion lake and a few colleagues want to read from
it. You attach the lake as usual, then start a server.

```{r build-lake}
attach_ducklake("trial", lake_path = "~/lakes/trial")

with_transaction(
  create_table(mtcars, "vehicles"),
  author = "You",
  commit_message = "Seed the shared lake"
)
```

```{r serve}
install_quack()
quack_serve(token = "super_secret")
#> Quack server listening on "quack:localhost".
```

The server runs in the background of the DuckDB instance, so your R session stays
free for other work. Anyone who can reach the address can now connect with the
client code from the previous section, pointing at your machine and using the
same token.

Choose the address to match who should reach the server. `"quack:localhost"`
(the default) stays on your own machine, which is useful for trying things out.
To accept connections from other machines, listen on a host name or address they
can route to.

When you are finished, stop the server:

```{r stop}
quack_stop()
#> Quack server on "quack:localhost" stopped.
```

### A note on access

A token is the only thing standing between your data and anyone who can reach the
server. Pick a strong one. If you start a server without a token, `quack_serve()`
warns you, because an open server lets any client that can reach it read and
write the lake. On an untrusted network, put the server behind TLS or a reverse
proxy rather than exposing it directly. The DuckDB
[Quack documentation](https://duckdb.org/docs/current/quack/overview) covers
these deployment options.

## Quack or a shared catalog?

ducklake already supports PostgreSQL and SQLite catalogs for concurrent access
(see `?attach_ducklake` and the package NEWS). Quack covers a similar need from a
different angle, and the right choice depends on where your data lives and who
can reach it.

| | Shared catalog (PostgreSQL or SQLite) | Quack |
|---|---|---|
| What clients connect to | The catalog database and the Parquet storage directly | One DuckDB server |
| Storage access each client needs | Read and write access to the same data path | None; only the server touches the files |
| Extra service to run | A PostgreSQL server (SQLite needs none) | A DuckDB server process |
| Good fit when | Everyone can mount the same shared storage | The data sits on one machine and you want to share access to it |

A common pattern is to use both: a PostgreSQL catalog so the lake itself supports
many writers, and a Quack server in front so analysts who cannot mount the
storage can still connect.

## Caveats

Quack is in beta. The DuckDB team expects to settle the protocol, function names,
and default settings before a stable release in DuckDB 2.0, so pin your versions
if you depend on it in a pipeline. It needs DuckDB 1.5.3 or newer on both the
server and the client; ducklake's Quack functions check this and stop with a
clear message on older engines. For very high write volumes a purpose built
database is still the better tool, but for sharing a study lake across a small
team Quack is a light way to get everyone onto the same versioned data.

## Learn more

- [Quack overview](https://duckdb.org/docs/current/quack/overview), the DuckDB documentation.
- `?quack_serve`, `?attach_quack`, and `?quack_query` for the function reference.
- The [Storage and Backup Management](storage-and-backups.html) vignette for how a DuckLake is laid out on disk.
