---
title: "Working with Transactions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with Transactions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
knitr::opts_knit$set(root.dir = tempdir())
```

```{r setup, message=FALSE, eval=FALSE}
library(ducklake)
library(dplyr)
library(duckplyr)

# Setup for examples
install_ducklake()
attach_ducklake("my_ducklake")
create_table("https://blobs.duckdb.org/nl_stations.csv", "nl_train_stations")
```

## Transaction support

Group multiple operations together with ACID transactions. This ensures that either all operations succeed or none do, maintaining data consistency.

## Basic transaction workflow

```{r check-data, eval=FALSE}
# Check what data we currently have
get_ducklake_table("nl_train_stations") |>
  filter(code %in% c("HT", "ASB")) |>
  select(code, name_short) |>
  collect()
```

```{r commit-example, eval=FALSE}
# Start a transaction
begin_transaction()

# Make multiple changes atomically within the transaction
duckplyr::db_exec("UPDATE nl_train_stations SET name_short = 'COMMITTED_CHANGE' WHERE code = 'HT'")
duckplyr::db_exec("UPDATE nl_train_stations SET name_short = 'ALSO_COMMITTED' WHERE code = 'ASB'")

# Commit both changes together
commit_transaction()

# Add author and commit message metadata to the snapshot
set_snapshot_metadata(
  ducklake_name = "my_ducklake",
  author = "Data Team",
  commit_message = "Updated station names for clarity"
)

# Verify the changes were applied
get_ducklake_table("nl_train_stations") |>
  filter(code %in% c("HT", "ASB")) |>
  select(code, name_short) |>
  collect()
```

## View commit history

```{r view-history, eval=FALSE}
# View the recent commit history with metadata
get_metadata_table("ducklake_snapshot_changes", ducklake_name = "my_ducklake") |>
  select(snapshot_id, changes_made, author, commit_message) |>
  collect() |>
  tail(3)
```

## Rolling back transactions

If you need to undo changes before committing:

```{r rollback-example, eval=FALSE}
# Start another transaction
begin_transaction()

# Make a change we'll roll back
duckplyr::db_exec("UPDATE nl_train_stations SET name_short = 'ROLLBACK_TEST' WHERE code = 'HT'")

# Decide to rollback instead of commit
rollback_transaction()

# Verify the change was NOT applied (should still be "COMMITTED_CHANGE")
get_ducklake_table("nl_train_stations") |>
  filter(code == "HT") |>
  select(code, name_short) |>
  collect()
```

## Key concepts

- **`begin_transaction()`**: Start a new transaction
- **`commit_transaction()`**: Apply all changes made within the transaction
- **`rollback_transaction()`**: Discard all changes made within the transaction
- **`set_snapshot_metadata()`**: Add metadata (author, commit message) to snapshots

Transactions are essential for maintaining data integrity when making multiple related changes.

```{r cleanup, include=FALSE, eval=FALSE}
detach_ducklake("my_ducklake")
```
