---
title: "Visualizing Your Lake"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Visualizing Your Lake}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
# Evaluate chunks only where the ducklake DuckDB extension is already
# installed. The probe never downloads anything, so building this vignette
# needs no network access.
ducklake_available <- ducklake::ducklake_extension_available()
viz_deps <- ducklake_available && requireNamespace("ggplot2", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4,
  eval = viz_deps
)
# Use a unique temp directory for this vignette to avoid conflicts during R CMD check
vignette_temp_dir <- file.path(tempdir(), "visualizing_your_lake_vignette")
dir.create(vignette_temp_dir, showWarnings = FALSE, recursive = TRUE)
knitr::opts_knit$set(root.dir = vignette_temp_dir)
```

```{r setup, message=FALSE}
library(ducklake)
library(dplyr)
```

## Introduction

A DuckLake keeps detailed records about itself: every change to a table is a
snapshot, every changed row is in the change feed, and every Parquet file is
tracked in the catalog. All of that comes back to R as data frames, but past
a handful of rows a picture is easier to scan. This vignette walks through
the package's three plotting functions and its interactive viewer:

- `plot_snapshots()` draws a table's history as a commit-log style timeline
- `plot_table_changes()` shows how many rows each snapshot changed
- `view_table_changes()` opens the changed rows themselves in an
  interactive viewer
- `plot_table_files()` shows how each table's data is laid out on disk

The plots need the ggplot2 package and the viewer needs htmlwidgets;
ducklake suggests both but requires neither.

## Building Some History

We'll create a small lake and put a table through a few changes, recording
authors and commit messages along the way.

```{r create-datalake}
attach_ducklake(
  ducklake_name = "lake_viz_demo",
  lake_path = vignette_temp_dir,
  author = "Data Engineer"
)

fleet <- tibble(
  car_id = 1:5,
  model = c("Corolla", "Civic", "Model 3", "F-150", "Outback"),
  mileage = c(42000, 38000, 12000, 67000, 55000)
)

with_transaction(
  create_table(fleet, "fleet"),
  author = "Data Engineer",
  commit_message = "Initial fleet load"
)
```

Now a couple of routine updates:

```{r make-changes}
new_cars <- tibble(
  car_id = 6:7,
  model = c("CX-5", "Ioniq 5"),
  mileage = c(21000, 8000)
)

with_transaction(
  rows_insert(get_ducklake_table("fleet"), new_cars, by = "car_id"),
  author = "Fleet Manager",
  commit_message = "Add March arrivals"
)

serviced <- tibble(
  car_id = c(1, 3, 5),
  mileage = c(42750, 13400, 55900)
)

with_transaction(
  rows_update(get_ducklake_table("fleet"), serviced, by = "car_id"),
  author = "Fleet Manager",
  commit_message = "Record spring odometer readings"
)

sold <- tibble(car_id = 4L)

with_transaction(
  rows_delete(get_ducklake_table("fleet"), sold, by = "car_id"),
  author = "Fleet Manager",
  commit_message = "Remove sold F-150"
)
```

The audit trail so far:

```{r list-snapshots}
list_table_snapshots("fleet") |>
  select(snapshot_id, snapshot_time, author, commit_message)
```

## Plotting the Timeline

`plot_snapshots()` turns that data frame into a picture:

```{r plot-table}
plot_snapshots("fleet")
```

Reading it like a git log: the newest snapshot is at the top, each row shows
the snapshot's timestamp, author, and commit message, and the color shows
whether the snapshot created the table, changed data, changed the schema, or
was maintenance activity (like flushing inlined data or compacting files).
Snapshots are spaced evenly by order rather than by clock time, so a table
that sat untouched for months stays readable: a long idle stretch shows up
as an inline marker (like "103 days later") instead of stretching the whole
plot.

## How Much Changed

The timeline shows when changes happened and what kind they were;
`plot_table_changes()` shows how big each one was. It counts the rows each
snapshot touched using DuckLake's change feed (the same data
`get_table_changes()` returns) and draws them as diverging bars: rows
inserted or updated above the axis, rows deleted below it.

```{r plot-changes}
plot_table_changes("fleet")
```

## Browsing the Changes

`plot_table_changes()` counts the rows; `view_table_changes()` shows them.
It takes the change feed from `get_table_changes()` and opens it in an
interactive viewer. The sidebar lists the snapshots in range, each with
its author and message, the columns that changed, and the rows inserted,
updated, and deleted; the main panel shows whichever part you pick. An
update is one row, shown from its old or its new side, with the changed
cells highlighted and both values on hover.

```{r view-changes, eval = viz_deps && requireNamespace("htmlwidgets", quietly = TRUE)}
view_table_changes(get_table_changes("fleet"))
```

The feed is a lazy table, so dplyr verbs narrow it before anything is
collected. One car's history:

```{r view-changes-filter, eval = viz_deps && requireNamespace("htmlwidgets", quietly = TRUE)}
get_table_changes("fleet") |>
  filter(car_id == 3) |>
  view_table_changes()
```

A filter on a column whose value changed, like `mileage`, matches one
image of an update and not the other; the viewer marks such rows as
one-sided. Filters on `car_id`, `snapshot_id`, `rowid`, or `change_type`
keep pairs whole. The layout follows Hadley Wickham's
[data-diff](https://github.com/hadley/data-diff).

## How the Data Is Stored

Behind the scenes, table data lives in Parquet files (very small writes may
be inlined into the catalog until they are flushed out). To make the storage
view interesting, let's add a larger table alongside `fleet` and grow it in
two batches, so it ends up with two data files:

```{r create-telemetry}
set.seed(42)

telemetry <- tibble(
  reading_id = 1:5000,
  car_id = sample(1:7, 5000, replace = TRUE),
  speed = round(runif(5000, 0, 80)),
  fuel = round(runif(5000, 0, 100))
)

create_table(telemetry, "telemetry")

more_readings <- tibble(
  reading_id = 5001:10000,
  car_id = sample(1:7, 5000, replace = TRUE),
  speed = round(runif(5000, 0, 80)),
  fuel = round(runif(5000, 0, 100))
)

rows_insert(get_ducklake_table("telemetry"), more_readings, by = "reading_id")
```

The fleet table's handful of rows were small enough to be inlined, so we
flush them into a Parquet file to make them visible to the storage
statistics:

```{r flush}
flush_inlined_data()
```

`get_table_info()` reports each table's file count and size:

```{r table-info}
get_table_info()
```

And `plot_table_files()` draws it, with delete files (rows removed but not
yet compacted away) stacked in a second color:

```{r plot-files}
plot_table_files()
```

This is the picture to watch as a lake ages: many small files or a growing
delete-file share mean it is time for the compaction tools covered in
`vignette("storage-and-backups")`, like `merge_adjacent_files()` and
`rewrite_data_files()`.

## Plotting the Whole Lake

With two tables in the lake, we can step back and look at everything at
once. Calling `plot_snapshots()` without a table name switches to a swimlane
view: one row per table, one point per snapshot, with recently active tables
at the top. Where the single-table timeline shows what happened to one
table, the swimlane shows which tables are active and what kind of churn
each one sees:

```{r plot-lake}
plot_snapshots()
```

Snapshots that don't belong to any table, like the initial schema creation,
appear in a `(lake)` lane. As in the timeline, points are spaced by order
rather than clock time; the x axis labels show each snapshot's date.

## Customizing the Plots

Each function returns a regular ggplot object, so you can restyle it like
any other plot. (The single-table timeline draws on a blank canvas, so it is
best customized with `labs()` and `theme()` tweaks rather than a full theme
swap like `theme_classic()`, which would bring back axes that have no
meaning there.)

```{r plot-custom}
plot_snapshots("fleet") +
  ggplot2::labs(title = "Fleet table audit trail")
```

```{r cleanup, include = FALSE}
detach_ducklake("lake_viz_demo", shutdown = TRUE)
```

## Learn More

- `vignette("time-travel")` covers querying and restoring the historical
  versions these snapshots represent
- `vignette("transactions")` covers recording authors and commit messages
  with `with_transaction()` and `commit_transaction()`
- `vignette("storage-and-backups")` covers the maintenance functions that
  keep the file layout healthy
