---
title: "Visualizing Your Lake"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Visualizing Your Lake}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
# Skip if the duckdb R package is too old to support the ducklake v1.0 extension
duckdb_version_ok <- tryCatch({
  packageVersion("duckdb") >= "1.5.1"
}, error = function(e) FALSE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4,
  eval = duckdb_version_ok && requireNamespace("ggplot2", quietly = TRUE)
)
# Use a unique temp directory for this vignette to avoid conflicts during R CMD check
vignette_temp_dir <- file.path(tempdir(), "visualizing_your_lake_vignette")
dir.create(vignette_temp_dir, showWarnings = FALSE, recursive = TRUE)
knitr::opts_knit$set(root.dir = vignette_temp_dir)
```

```{r setup, message=FALSE}
library(ducklake)
library(dplyr)
```

## Introduction

A DuckLake keeps detailed records about itself: every change to a table is a
snapshot, every changed row is in the change feed, and every Parquet file is
tracked in the catalog. All of that comes back to R as data frames, but past
a handful of rows a picture is easier to scan. This vignette walks through
the package's three plotting functions:

- `plot_snapshots()` draws a table's history as a commit-log style timeline
- `plot_table_changes()` shows how many rows each snapshot changed
- `plot_table_files()` shows how each table's data is laid out on disk

These require the ggplot2 package, which is suggested (not required) by
ducklake.

## Building Some History

We'll create a small lake and put a table through a few changes, recording
authors and commit messages along the way.

```{r create-datalake}
install_ducklake()

attach_ducklake(
  ducklake_name = "lake_viz_demo",
  lake_path = vignette_temp_dir
)

fleet <- tibble(
  car_id = 1:5,
  model = c("Corolla", "Civic", "Model 3", "F-150", "Outback"),
  mileage = c(42000, 38000, 12000, 67000, 55000)
)

with_transaction(
  create_table(fleet, "fleet"),
  author = "Data Engineer",
  commit_message = "Initial fleet load"
)
```

Now a couple of routine updates:

```{r make-changes}
new_cars <- tibble(
  car_id = 6:7,
  model = c("CX-5", "Ioniq 5"),
  mileage = c(21000, 8000)
)

with_transaction(
  rows_insert(get_ducklake_table("fleet"), new_cars, by = "car_id"),
  author = "Fleet Manager",
  commit_message = "Add March arrivals"
)

serviced <- tibble(
  car_id = c(1, 3, 5),
  mileage = c(42750, 13400, 55900)
)

with_transaction(
  rows_update(get_ducklake_table("fleet"), serviced, by = "car_id"),
  author = "Fleet Manager",
  commit_message = "Record spring odometer readings"
)

sold <- tibble(car_id = 4L)

with_transaction(
  rows_delete(get_ducklake_table("fleet"), sold, by = "car_id"),
  author = "Fleet Manager",
  commit_message = "Remove sold F-150"
)
```

The audit trail so far:

```{r list-snapshots}
list_table_snapshots("fleet") |>
  select(snapshot_id, snapshot_time, author, commit_message)
```

## Plotting the Timeline

`plot_snapshots()` turns that data frame into a picture:

```{r plot-table}
plot_snapshots("fleet")
```

Reading it like a git log: the newest snapshot is at the top, the horizontal
position shows when each change landed, and the color shows whether the
snapshot created the table, changed data, changed the schema, or was
maintenance activity (like flushing inlined data or compacting files).

## How Much Changed

The timeline shows when changes happened and what kind they were;
`plot_table_changes()` shows how big each one was. It counts the rows each
snapshot touched using DuckLake's change feed (the same data
`get_table_changes()` returns) and draws them as diverging bars: rows
inserted or updated above the axis, rows deleted below it.

```{r plot-changes}
plot_table_changes("fleet")
```

## Plotting the Whole Lake

Calling `plot_snapshots()` without a table name shows every snapshot in the
lake, which is useful for a quick health check of overall activity:

```{r plot-lake}
plot_snapshots()
```

## How the Data Is Stored

Behind the scenes, table data lives in Parquet files (very small writes may
be inlined into the catalog until they are flushed out). To make the storage
view interesting, let's add a larger table alongside `fleet` and grow it in
two batches, so it ends up with two data files:

```{r create-telemetry}
set.seed(42)

telemetry <- tibble(
  reading_id = 1:5000,
  car_id = sample(1:7, 5000, replace = TRUE),
  speed = round(runif(5000, 0, 80)),
  fuel = round(runif(5000, 0, 100))
)

create_table(telemetry, "telemetry")

more_readings <- tibble(
  reading_id = 5001:10000,
  car_id = sample(1:7, 5000, replace = TRUE),
  speed = round(runif(5000, 0, 80)),
  fuel = round(runif(5000, 0, 100))
)

rows_insert(get_ducklake_table("telemetry"), more_readings, by = "reading_id")
```

The fleet table's handful of rows were small enough to be inlined, so we
flush them into a Parquet file to make them visible to the storage
statistics:

```{r flush}
flush_inlined_data()
```

`get_table_info()` reports each table's file count and size:

```{r table-info}
get_table_info()
```

And `plot_table_files()` draws it, with delete files (rows removed but not
yet compacted away) stacked in a second color:

```{r plot-files}
plot_table_files()
```

This is the picture to watch as a lake ages: many small files or a growing
delete-file share mean it is time for the compaction tools covered in
`vignette("storage-and-backups")`, like `merge_adjacent_files()` and
`rewrite_data_files()`.

## Customizing the Plots

Each function returns a regular ggplot object, so you can restyle it like
any other plot:

```{r plot-custom}
plot_snapshots("fleet") +
  ggplot2::theme_classic() +
  ggplot2::labs(title = "Fleet table audit trail")
```

```{r cleanup, include = FALSE}
detach_ducklake("lake_viz_demo", shutdown = TRUE)
```

## Learn More

- `vignette("time-travel")` covers querying and restoring the historical
  versions these snapshots represent
- `vignette("transactions")` covers recording authors and commit messages
  with `with_transaction()` and `commit_transaction()`
- `vignette("storage-and-backups")` covers the maintenance functions that
  keep the file layout healthy
