---
title: "Visualizing Snapshot History"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Visualizing Snapshot History}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
# Skip if the duckdb R package is too old to support the ducklake v1.0 extension
duckdb_version_ok <- tryCatch({
  packageVersion("duckdb") >= "1.5.1"
}, error = function(e) FALSE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4,
  eval = duckdb_version_ok && requireNamespace("ggplot2", quietly = TRUE)
)
# Use a unique temp directory for this vignette to avoid conflicts during R CMD check
vignette_temp_dir <- file.path(tempdir(), "visualizing_snapshots_vignette")
dir.create(vignette_temp_dir, showWarnings = FALSE, recursive = TRUE)
knitr::opts_knit$set(root.dir = vignette_temp_dir)
```

```{r setup, message=FALSE}
library(ducklake)
library(dplyr)
```

## Introduction

Every change to a DuckLake table is recorded as a snapshot, and
`list_table_snapshots()` returns that history as a data frame. For a table
that has been through more than a handful of changes, a picture is easier to
scan than the raw rows. `plot_snapshots()` draws the history as a commit-log
style timeline: one row per snapshot, positioned by when it happened, colored
by what kind of change it was, and annotated with the author and commit
message when those were recorded.

This requires the ggplot2 package, which is suggested (not required) by
ducklake.

## Building Some History

We'll create a small lake and put a table through a few changes, recording
authors and commit messages along the way.

```{r create-datalake}
install_ducklake()

attach_ducklake(
  ducklake_name = "snapshot_viz_demo",
  lake_path = vignette_temp_dir
)

fleet <- tibble(
  car_id = 1:5,
  model = c("Corolla", "Civic", "Model 3", "F-150", "Outback"),
  mileage = c(42000, 38000, 12000, 67000, 55000)
)

with_transaction(
  create_table(fleet, "fleet"),
  author = "Data Engineer",
  commit_message = "Initial fleet load"
)
```

Now a couple of routine updates:

```{r make-changes}
new_cars <- tibble(
  car_id = 6:7,
  model = c("CX-5", "Ioniq 5"),
  mileage = c(21000, 8000)
)

with_transaction(
  rows_insert(get_ducklake_table("fleet"), new_cars, by = "car_id"),
  author = "Fleet Manager",
  commit_message = "Add March arrivals"
)

sold <- tibble(car_id = 4L)

with_transaction(
  rows_delete(get_ducklake_table("fleet"), sold, by = "car_id"),
  author = "Fleet Manager",
  commit_message = "Remove sold F-150"
)
```

The audit trail so far:

```{r list-snapshots}
list_table_snapshots("fleet") |>
  select(snapshot_id, snapshot_time, author, commit_message)
```

## Plotting the Timeline

`plot_snapshots()` turns that data frame into a picture:

```{r plot-table}
plot_snapshots("fleet")
```

Reading it like a git log: the newest snapshot is at the top, the horizontal
position shows when each change landed, and the color shows whether the
snapshot created the table, changed data, changed the schema, or was
maintenance activity (like flushing inlined data or compacting files).

## Customizing the Plot

The function returns a regular ggplot object, so you can restyle it like any
other plot:

```{r plot-custom}
plot_snapshots("fleet") +
  ggplot2::theme_classic() +
  ggplot2::labs(title = "Fleet table audit trail")
```

## Plotting the Whole Lake

Calling `plot_snapshots()` without a table name shows every snapshot in the
lake, which is useful for a quick health check of overall activity:

```{r plot-lake}
plot_snapshots()
```

```{r cleanup, include = FALSE}
detach_ducklake("snapshot_viz_demo", shutdown = TRUE)
```

## Learn More

- `vignette("time-travel")` covers querying and restoring the historical
  versions these snapshots represent
- `vignette("transactions")` covers recording authors and commit messages
  with `with_transaction()` and `commit_transaction()`
