---
title: "Clinical Trial Data Lake with ducklake"
---

```{r, include = FALSE}
# Evaluate chunks only where the ducklake DuckDB extension is already
# installed (the probe never downloads anything) and the packages this
# article needs are present. The pkgdown workflow installs all of them.
has_deps <- all(vapply(
  c("pharmaversesdtm", "admiral", "haven", "labelled", "ggplot2"),
  requireNamespace,
  logical(1),
  quietly = TRUE
))
ducklake_available <- ducklake::ducklake_extension_available()
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = has_deps && ducklake_available,
  fig.width = 7,
  fig.height = 4.5
)
# A directory of its own keeps this lake and its XPT transfer apart from
# other articles built in the same session. Start it empty, so a lake left
# behind by an earlier knit cannot make the schema creation below fail.
article_dir <- file.path(tempdir(), "clinical_trial_article")
unlink(article_dir, recursive = TRUE)
dir.create(article_dir, recursive = TRUE)
knitr::opts_knit$set(root.dir = article_dir)
```

```{r setup, message = FALSE}
library(ducklake)
library(dplyr)
library(purrr)
library(admiral)
library(stringr)
```

An EDC system logs every change to the data it holds. Once SDTM datasets
leave it for statistical programming, that protection stops. Domains
travel as XPT files, derived datasets pile up as copies in folders, and
the questions a reviewer asks later get answered from memory and file
names: which version of ADAE fed this table, what changed between two data
cuts, who corrected that record and why.

This article builds a clinical trial data lake with ducklake that keeps
those answers in the catalog. It loads an SDTM transfer as received,
standardizes it, derives ADSL and ADAE with
[admiral](https://pharmaverse.github.io/admiral/), stores an analysis
result next to the datasets that produced it, and then does the things
that happen after the datasets exist: adds a derived variable, pushes a
correction from SDTM through ADaM to the result, and reads an earlier
version back. Every write is a commit with an author and a message, and
the history is one query away. The data is the CDISC pilot study as
shipped by [pharmaversesdtm](https://pharmaverse.github.io/pharmaversesdtm/).

## A lake with four layers

`attach_ducklake()` opens a lake, creating it first when nothing is at the
path yet. The article uses a temporary directory; a study would use a
project directory or a shared location, and
[Choosing a Deployment](deployment.html) covers the catalog options for a
team.

```{r attach, message = FALSE}
attach_ducklake("clinical_trial_lake", lake_path = article_dir)
```

The lake follows the medallion pattern, one schema per layer, so a table's
name says how far from the source it is and access can be granted layer by
layer:

```
bronze    SDTM as received: XPT files, blanks for missing values, labels
silver    SDTM standardized: blanks are NA, labels live in the catalog
gold      ADaM datasets, derived with admiral
results   analysis results, derived from gold
```

```{r schemas}
with_transaction({
  create_schema("bronze")
  create_schema("silver")
  create_schema("gold")
  create_schema("results")
}, author = "Data Manager", commit_message = "Create the lake's layers")
```

Every write to a lake is a commit. `with_transaction()` records who made it
and why, lands everything inside it as one snapshot, and rolls the whole
thing back if any step fails. The confirmation names the snapshot, and
those numbers are what the history section reads back at the end.

## Bronze: the SDTM transfer

SDTM reaches the statistical programmers as SAS transport (XPT) files, the
format FDA accepts. pharmaversesdtm ships the pilot domains as data frames,
so the article writes four of them out as XPT first, to stand in for a
transfer from data management. Version 5 of the format caps dataset and
variable names at eight characters, which is one reason SDTM and ADaM
names look the way they do.

```{r transfer}
sdtm <- list(
  dm = pharmaversesdtm::dm,
  ds = pharmaversesdtm::ds,
  ex = pharmaversesdtm::ex,
  ae = pharmaversesdtm::ae
)

sdtm_transfer <- file.path(article_dir, "sdtm_transfer")
dir.create(sdtm_transfer, showWarnings = FALSE)
iwalk(sdtm, \(data, domain) {
  haven::write_xpt(
    data,
    file.path(sdtm_transfer, paste0(domain, ".xpt")),
    version = 5,
    name = domain
  )
})
```

The bronze layer keeps the transfer exactly as it arrived, so the lake can
always show what data management sent and reprocess it when the cleaning
logic changes. One commit loads the four domains:

```{r bronze}
with_transaction({
  walk(names(sdtm), \(domain) {
    create_table(
      haven::read_xpt(file.path(sdtm_transfer, paste0(domain, ".xpt"))),
      paste0("bronze.", domain)
    )
  })
}, author = "Data Manager", commit_message = "Load the SDTM transfer as received")
```

Each `create_table()` reported the labels it stored. haven reads the XPT
variable labels into `label` attributes, and the lake keeps them as column
comments, where any client of the catalog can read them:

```{r bronze-labels}
get_table_comments("bronze.dm") |>
  filter(object_type == "column") |>
  select(column_name, comment) |>
  head(5)
```

XPT also preserves the SDTM convention for a missing character value,
which is a blank. `DTHDTC` holds the date of death, so it is blank for most
subjects:

```{r bronze-blanks}
get_ducklake_table("bronze.dm") |>
  count(dthdtc_blank = DTHDTC == "") |>
  collect()
```

## Silver: blanks to NA, inside DuckDB

R analysis code, admiral included, expects `NA` where SAS had a blank.
admiral's `convert_blanks_to_na()` does the conversion on a data frame. On
a lazy lake table it would do nothing, because nothing has been read into
R yet, so the silver layer writes the same conversion as a dplyr pipeline:
dbplyr translates `na_if()` into `NULLIF()`, and `create_table()` runs the
whole query inside DuckDB and writes the result straight into the lake. A
silver table is derived without its rows entering R.

```{r blanks-helper}
# The in-database counterpart of admiral::convert_blanks_to_na(). A lazy
# table cannot inspect its own types, so a zero-row read supplies the
# character column names.
blanks_to_na <- function(tbl) {
  types <- tbl |> head(0) |> collect()
  chr_cols <- names(select(types, where(is.character)))
  tbl |> mutate(across(all_of(chr_cols), ~ na_if(.x, "")))
}
```

Each silver recipe is kept in a named list, under the name it is
materialized as, because the recipe that builds a layer is also its
lineage, which the results section comes back to. `iwalk()` then hands
each recipe and its name to `create_table()`.

```{r silver}
silver_recipes <- list(
  "silver.dm" = get_ducklake_table("bronze.dm") |> blanks_to_na(),
  "silver.ds" = get_ducklake_table("bronze.ds") |> blanks_to_na(),
  "silver.ex" = get_ducklake_table("bronze.ex") |> blanks_to_na(),
  "silver.ae" = get_ducklake_table("bronze.ae") |> blanks_to_na()
)

with_transaction(
  iwalk(silver_recipes, create_table),
  author = "Data Manager",
  commit_message = "Standardize SDTM: blanks to NA"
)

get_ducklake_table("silver.dm") |>
  count(dthdtc_missing = is.na(DTHDTC)) |>
  collect()
```

The labels followed the columns: a table derived from a lake table keeps
the comments of the columns it selects, so the silver layer is documented
without another step. [Views, Comments, and Labels](views-comments-labels.html)
covers what else lives in the catalog.

## Gold: ADaM datasets with admiral

admiral works on data frames, so the gold layer starts by collecting the
silver tables. `collect()` reattaches the stored comments as `label`
attributes, and the frames look the way they would have coming straight
from haven, labels and all.

```{r collect-silver}
dm <- get_ducklake_table("silver.dm") |> collect()
ds <- get_ducklake_table("silver.ds") |> collect()
ex <- get_ducklake_table("silver.ex") |> collect()
ae <- get_ducklake_table("silver.ae") |> collect()

attr(dm$AGE, "label")
```

### ADSL

ADSL holds one record per subject: treatment, treatment dates, population
flags, disposition, and the groupings the analyses use. The derivation
below follows admiral's ADSL template (`admiral::use_ad_template("adsl")`)
and keeps the parts a safety analysis needs.

```{r build-adsl}
# End of study status from the disposition event, in the SBJTSTAT terms
format_eosstt <- function(x) {
  case_when(
    x == "COMPLETED" ~ "COMPLETED",
    x == "SCREEN FAILURE" ~ NA_character_,
    !is.na(x) ~ "DISCONTINUED",
    TRUE ~ "ONGOING"
  )
}

# One lookup derives the character and numeric versions of a grouping
agegr1_lookup <- exprs(
  ~condition,           ~AGEGR1, ~AGEGR1N,
  AGE < 18,               "<18",        1,
  between(AGE, 18, 64), "18-64",        2,
  AGE > 64,               ">64",        3,
  is.na(AGE),         "Missing",        4
)

# Exposure datetimes: impute start times to the first and end times to the
# last moment of the day, and keep the flags that say so
ex_ext <- ex |>
  derive_vars_dtm(dtc = EXSTDTC, new_vars_prefix = "EXST") |>
  derive_vars_dtm(dtc = EXENDTC, new_vars_prefix = "EXEN", time_imputation = "last")

ds_ext <- ds |>
  derive_vars_dt(dtc = DSSTDTC, new_vars_prefix = "DSST")

adsl <- dm |>
  # DOMAIN is an SDTM variable; ADSL has none
  select(-DOMAIN) |>
  mutate(TRT01P = ARM, TRT01A = ACTARM) |>
  # First and last exposure
  derive_vars_merged(
    dataset_add = ex_ext,
    filter_add = (EXDOSE > 0 | (EXDOSE == 0 & str_detect(EXTRT, "PLACEBO"))) &
      !is.na(EXSTDTM),
    new_vars = exprs(TRTSDTM = EXSTDTM, TRTSTMF = EXSTTMF),
    order = exprs(EXSTDTM, EXSEQ),
    mode = "first",
    by_vars = exprs(STUDYID, USUBJID)
  ) |>
  derive_vars_merged(
    dataset_add = ex_ext,
    filter_add = (EXDOSE > 0 | (EXDOSE == 0 & str_detect(EXTRT, "PLACEBO"))) &
      !is.na(EXENDTM),
    new_vars = exprs(TRTEDTM = EXENDTM, TRTETMF = EXENTMF),
    order = exprs(EXENDTM, EXSEQ),
    mode = "last",
    by_vars = exprs(STUDYID, USUBJID)
  ) |>
  derive_vars_dtm_to_dt(source_vars = exprs(TRTSDTM, TRTEDTM)) |>
  derive_var_trtdurd() |>
  # Disposition: randomization, end of study, death
  derive_vars_merged(
    dataset_add = ds_ext,
    by_vars = exprs(STUDYID, USUBJID),
    new_vars = exprs(RANDDT = DSSTDT),
    filter_add = DSDECOD == "RANDOMIZED"
  ) |>
  derive_vars_merged(
    dataset_add = ds_ext,
    by_vars = exprs(STUDYID, USUBJID),
    new_vars = exprs(EOSDT = DSSTDT),
    filter_add = DSCAT == "DISPOSITION EVENT" & DSDECOD != "SCREEN FAILURE"
  ) |>
  derive_vars_merged(
    dataset_add = ds_ext,
    by_vars = exprs(STUDYID, USUBJID),
    filter_add = DSCAT == "DISPOSITION EVENT",
    new_vars = exprs(EOSSTT = format_eosstt(DSDECOD)),
    missing_values = exprs(EOSSTT = "ONGOING")
  ) |>
  derive_vars_dt(
    new_vars_prefix = "DTH",
    dtc = DTHDTC,
    highest_imputation = "M",
    date_imputation = "first"
  ) |>
  # Population flags are Y or N, never blank
  derive_var_merged_exist_flag(
    dataset_add = ex,
    by_vars = exprs(STUDYID, USUBJID),
    new_var = SAFFL,
    false_value = "N",
    missing_value = "N",
    condition = (EXDOSE > 0 | (EXDOSE == 0 & str_detect(EXTRT, "PLACEBO")))
  ) |>
  derive_var_merged_exist_flag(
    dataset_add = ds,
    by_vars = exprs(STUDYID, USUBJID),
    new_var = RANDFL,
    false_value = "N",
    missing_value = "N",
    condition = DSDECOD == "RANDOMIZED"
  ) |>
  derive_vars_cat(definition = agegr1_lookup)
```

The DM variables kept their SDTM labels through the pipeline. The derived
variables are new, so they get theirs here, in the standard ADaM wording;
`create_table()` stores every `label` attribute it finds, and the dataset
label goes on the table itself.

```{r adsl-labels}
adsl_labels <- list(
  TRT01P = "Planned Treatment for Period 01",
  TRT01A = "Actual Treatment for Period 01",
  TRTSDTM = "Datetime of First Exposure to Treatment",
  TRTSTMF = "Time of First Exposure Imput. Flag",
  TRTEDTM = "Datetime of Last Exposure to Treatment",
  TRTETMF = "Time of Last Exposure Imput. Flag",
  TRTSDT = "Date of First Exposure to Treatment",
  TRTEDT = "Date of Last Exposure to Treatment",
  TRTDURD = "Total Treatment Duration (Days)",
  RANDDT = "Date of Randomization",
  EOSDT = "End of Study Date",
  EOSSTT = "End of Study Status",
  DTHDT = "Date of Death",
  DTHDTF = "Date of Death Imputation Flag",
  SAFFL = "Safety Population Flag",
  RANDFL = "Randomized Population Flag",
  AGEGR1 = "Pooled Age Group 1",
  AGEGR1N = "Pooled Age Group 1 (N)"
)
adsl <- labelled::set_variable_labels(adsl, .labels = adsl_labels)

with_transaction({
  create_table(adsl, "gold.adsl")
  set_table_comment("gold.adsl", "Subject-Level Analysis Dataset")
}, author = "Statistical Programmer", commit_message = "Build ADSL")
```

```{r adsl-preview}
get_ducklake_table("gold.adsl") |>
  select(USUBJID, TRT01A, TRTSDT, TRTDURD, SAFFL, EOSSTT, AGEGR1) |>
  arrange(USUBJID) |>
  head(5) |>
  collect()

get_ducklake_table("gold.adsl") |>
  count(EOSSTT) |>
  arrange(EOSSTT) |>
  collect()
```

Screen failures have no end-of-study status, and the population flags say
which subjects the safety analyses count.

### ADAE

ADAE is an occurrence dataset: one record per adverse event, with analysis
dates, the treatment-emergent flag, and the first-occurrence flag the
incidence tables need. The derivation is a function, because the
correction section rebuilds the dataset with it. It follows admiral's ADAE
template.

```{r build-adae}
adsl_vars <- exprs(TRTSDT, TRTEDT, DTHDT, EOSDT)

adae_labels <- list(
  ASTDTM = "Analysis Start Datetime",
  ASTDTF = "Analysis Start Date Imputation Flag",
  ASTTMF = "Analysis Start Time Imputation Flag",
  AENDTM = "Analysis End Datetime",
  AENDTF = "Analysis End Date Imputation Flag",
  AENTMF = "Analysis End Time Imputation Flag",
  ASTDT = "Analysis Start Date",
  AENDT = "Analysis End Date",
  ASTDY = "Analysis Start Relative Day",
  AENDY = "Analysis End Relative Day",
  ADURN = "Analysis Duration (N)",
  ADURU = "Analysis Duration Units",
  ASEV = "Analysis Severity/Intensity",
  AREL = "Analysis Causality",
  TRTEMFL = "Treatment Emergent Analysis Flag",
  AOCCPFL = "1st Occurrence of Preferred Term Flag"
)

build_adae <- function(ae, adsl) {
  ae |>
    derive_vars_merged(
      dataset_add = adsl,
      new_vars = adsl_vars,
      by_vars = exprs(STUDYID, USUBJID)
    ) |>
    # Analysis dates: partial dates are imputed, and the flags record how.
    # A start date cannot fall before first dose, nor an end date after
    # death or the end of study
    derive_vars_dtm(
      dtc = AESTDTC,
      new_vars_prefix = "AST",
      highest_imputation = "M",
      min_dates = exprs(TRTSDT)
    ) |>
    derive_vars_dtm(
      dtc = AEENDTC,
      new_vars_prefix = "AEN",
      highest_imputation = "M",
      date_imputation = "last",
      time_imputation = "last",
      max_dates = exprs(DTHDT, EOSDT)
    ) |>
    derive_vars_dtm_to_dt(exprs(ASTDTM, AENDTM)) |>
    derive_vars_dy(reference_date = TRTSDT, source_vars = exprs(ASTDT, AENDT)) |>
    derive_vars_duration(
      new_var = ADURN,
      new_var_unit = ADURU,
      start_date = ASTDT,
      end_date = AENDT
    ) |>
    mutate(ASEV = AESEV, AREL = AEREL) |>
    # Treatment emergent: on or after first dose, up to 30 days after the
    # last one
    derive_var_trtemfl(
      trt_start_date = TRTSDT,
      trt_end_date = TRTEDT,
      end_window = 30
    ) |>
    # First occurrence of each preferred term, among treatment-emergent
    # events only
    restrict_derivation(
      derivation = derive_var_extreme_flag,
      args = params(
        by_vars = exprs(USUBJID, AEDECOD),
        order = exprs(ASTDTM, AESEQ),
        new_var = AOCCPFL,
        mode = "first"
      ),
      filter = TRTEMFL == "Y"
    ) |>
    # The rest of ADSL, so the dataset analyzes on its own
    derive_vars_merged(
      dataset_add = select(adsl, !!!negate_vars(adsl_vars)),
      by_vars = exprs(STUDYID, USUBJID)
    ) |>
    # restrict_derivation() rebinds the rows and drops label attributes on
    # the way, so the SDTM and ADSL labels are copied back
    labelled::copy_labels_from(ae) |>
    labelled::copy_labels_from(adsl) |>
    labelled::set_variable_labels(.labels = adae_labels)
}

adae <- build_adae(ae, adsl)

with_transaction({
  create_table(adae, "gold.adae")
  set_table_comment("gold.adae", "Adverse Events Analysis Dataset")
}, author = "Statistical Programmer", commit_message = "Build ADAE")
```

```{r adae-preview}
get_ducklake_table("gold.adae") |>
  filter(TRTEMFL == "Y") |>
  arrange(USUBJID, ASTDT, AESEQ) |>
  select(USUBJID, AEDECOD, ASTDT, ASTDTF, ASEV, AOCCPFL) |>
  head(5) |>
  collect()
```

The other analysis datasets a study needs come from the same templates and
land in the lake the same way: build the data frame with admiral, label
it, and `create_table()` it into `gold` inside a commit.

## Results: an analysis result stored with its inputs

A treatment-emergent adverse event summary is a query over gold: events by
actual treatment and severity, with the safety population as the
denominator, which has to come from ADSL because subjects with no events
are not in ADAE. Written as a lazy pipeline, the summary runs inside
DuckDB and lands in the results layer as a table of its own, versioned
like everything else, with the commit saying which display it belongs to.

```{r results}
safety_n <- get_ducklake_table("gold.adsl") |>
  filter(SAFFL == "Y") |>
  group_by(TRT01A) |>
  summarise(n_saf = n_distinct(USUBJID), .groups = "drop")

teae_recipe <- get_ducklake_table("gold.adae") |>
  filter(TRTEMFL == "Y") |>
  group_by(TRT01A, ASEV) |>
  summarise(
    n_subjects = n_distinct(USUBJID),
    n_events = n(),
    .groups = "drop"
  ) |>
  left_join(safety_n, by = "TRT01A") |>
  mutate(pct_subjects = 100 * n_subjects / n_saf) |>
  arrange(TRT01A, ASEV)

with_transaction(
  create_table(teae_recipe, "results.teae_summary"),
  author = "Statistician",
  commit_message = "Table 14.3.1: TEAEs by treatment and severity"
)

get_ducklake_table("results.teae_summary") |> collect()
```

### Where the numbers came from

ducklake records lineage at the table level: which tables each snapshot
touched, and why. The companion package
[dplyneage](https://tgerke.github.io/dplyneage/) traces lineage within a
query, from each output column back to the source columns it came from.
The recipe that built the summary is a lazy table, which is what
`extract_lineage()` reads. Passing it under the name it was materialized
as draws the hop from gold to results, and the gold nodes show only the
columns the recipe touched.

```{r lineage, eval = has_deps && ducklake_available && requireNamespace("dplyneage", quietly = TRUE)}
library(dplyneage)

teae_lineage <- extract_lineage(list("results.teae_summary" = teae_recipe))
lineage_flow(teae_lineage, height = "450px")
```

In an R session the diagram is interactive: hover a column for its type
and label, click one to isolate what feeds it. The same lineage answers as
data:

```{r lineage-upstream, eval = has_deps && ducklake_available && requireNamespace("dplyneage", quietly = TRUE)}
lineage_upstream(teae_lineage, "results.teae_summary.pct_subjects")
```

The admiral layers are not in the picture, and cannot be: admiral runs on
data frames and leaves no query tree behind. Their traceability is the
metadata a submission carries anyway, since the ADaM Implementation Guide
(section 2.2) asks for the derivation of every analysis variable to be
described in metadata. So the division of labor is: lineage from the lake
wherever the pipeline is a query, ADaM metadata where the standard asks
for it. The silver recipes stitch into the same graph with
`extract_lineage(silver_recipes)`, and dplyneage's
[lineage that travels with the data](https://tgerke.github.io/dplyneage/articles/ducklake-versioned-lineage.html)
article stores each hop's lineage on the commit that wrote it, through
`commit_extra_info`, so a snapshot's rows and their derivation come back
together.

## Working with the lake

### Queries across layers

Lake tables are lazy: dplyr verbs become SQL that DuckDB runs, and only
the result reaches R. A check that every adverse event record belongs to a
subject in ADSL reads the silver and gold layers together without
collecting either:

```{r integrity}
get_ducklake_table("silver.ae") |>
  anti_join(get_ducklake_table("gold.adsl"), by = "USUBJID") |>
  count(name = "ae_records_without_adsl_subject") |>
  collect()
```

### A derived variable added in place

When a reviewer asks for the protocol's age grouping as well, the column
is declared, filled with an in-database `UPDATE`, and labelled, as one
commit. No rows pass through R, and the versions before and after are both
readable.

```{r agegr2}
with_transaction({
  add_table_column("gold.adsl", "AGEGR2", "VARCHAR")
  add_table_column("gold.adsl", "AGEGR2N", "INTEGER")
  get_ducklake_table("gold.adsl") |>
    mutate(
      AGEGR2 = case_when(AGE < 65 ~ "<65", AGE <= 80 ~ "65-80", TRUE ~ ">80"),
      AGEGR2N = case_when(AGE < 65 ~ 1L, AGE <= 80 ~ 2L, TRUE ~ 3L)
    ) |>
    ducklake_exec()
  set_column_comments(
    "gold.adsl",
    AGEGR2 = "Pooled Age Group 2",
    AGEGR2N = "Pooled Age Group 2 (N)"
  )
}, author = "Statistical Programmer", commit_message = "Add the protocol's age grouping, AGEGR2")

get_ducklake_table("gold.adsl") |>
  count(AGEGR2N, AGEGR2) |>
  arrange(AGEGR2N) |>
  collect()
```

[Modifying Tables](modifying-tables.html) covers the choice between
`ducklake_exec()`, the `rows_*()` functions, `merge_into()`, and
`replace_table()`.

### A correction that flows downstream

Data corrections do not start in ADaM. A query on an adverse event
resolves, the SDTM record changes, and everything derived from it is
rebuilt by running the same programs again. In the lake that is one
transaction: the SDTM correction as a single in-place update through
`rows_update()`, ADAE rebuilt by the same `build_adae()`, and the summary
rebuilt from its recipe. `replace_table()` rewrites a table as one
versioned change and carries its labels and settings over. Either all
three land as one snapshot, or none of them does.

```{r correction}
with_transaction({
  rows_update(
    get_ducklake_table("silver.ae"),
    data.frame(USUBJID = "01-701-1015", AESEQ = 1, AESEV = "SEVERE"),
    by = c("USUBJID", "AESEQ")
  )
  ae <- get_ducklake_table("silver.ae") |> collect()
  adsl <- get_ducklake_table("gold.adsl") |> collect()
  replace_table(build_adae(ae, adsl), "gold.adae")
  replace_table(teae_recipe, "results.teae_summary")
}, author = "Data Manager", commit_message = "DQ-042: AESEV corrected; ADAE and results rebuilt")
```

The change feed shows what the correction did to `silver.ae`, as the value
before and the value after:

```{r change-feed}
correction <- max(list_table_snapshots()$snapshot_id)

get_table_changes("silver.ae", correction, correction) |>
  select(change_type, USUBJID, AESEQ, AESEV) |>
  collect()
```

That is the shape of an audit trail as FDA describes it in its 2024
questions and answers on electronic systems in clinical investigations
(Q12 to Q14):
the date and time (`snapshot_time`), the individual (`author`), the old
and the new value (the change feed), and the reason (`commit_message`).
Prior entries are never obscured: a change never overwrites an earlier
version in place, and the rows before it stay readable until a retention
policy expires them. ICH E6(R3) (section 4.2.2) puts audit trails and
their review inside a risk-based data governance process. The lake does
not validate itself, though: qualifying the computing environment that
runs these programs stays with the sponsor.

## The history

`list_table_snapshots()` lists every commit in the lake, across all four
layers, with the author and message each one was given:

```{r history}
list_table_snapshots() |>
  select(snapshot_id, snapshot_time, author, commit_message)
```

`plot_snapshots()` draws one table's history as a timeline, with the
author and message of each commit. For ADAE, that is its build and the
rebuild the correction caused:

```{r plot-history, fig.width = 9, fig.height = 3}
plot_snapshots("gold.adae")
```

Any earlier version of a table can be read as a lazy table. ADAE before
the correction, next to ADAE now:

```{r time-travel}
adae_versions <- list_table_snapshots("gold.adae")
adae_versions |> select(snapshot_id, author, commit_message)

get_ducklake_table_version("gold.adae", min(adae_versions$snapshot_id)) |>
  filter(USUBJID == "01-701-1015", AESEQ == 1) |>
  select(USUBJID, AESEQ, ASEV) |>
  collect()

get_ducklake_table("gold.adae") |>
  filter(USUBJID == "01-701-1015", AESEQ == 1) |>
  select(USUBJID, AESEQ, ASEV) |>
  collect()
```

An analysis run against a data cut pins the whole session to a snapshot
with `attach_ducklake(snapshot_version = )`, so every table it reads is
the version of that cut. [Time Travel](time-travel.html) covers that,
along with reading as of a timestamp, comparing versions, and restoring
one.

## Before archival

Small changes such as the correction above are stored inline in the
catalog rather than as tiny Parquet files, which is what keeps a lake with
many small commits tidy. Before an archive or a transfer, flush them, so
the data path is a self-describing set of Parquet files that any tool can
read without ducklake, next to the catalog that holds the history.
[Data Inlining](data-inlining.html) has the details.

```{r flush}
flush_inlined_data()
```

## Detach

```{r detach}
detach_ducklake("clinical_trial_lake")
```

Detaching deletes nothing. Attaching the same path again picks up where
this session left off, history included.

## Learn more

- [Modifying Tables](modifying-tables.html): choosing how to change a table
- [Views, Comments, and Labels](views-comments-labels.html): shared query logic and documentation that live in the lake
- [Time Travel](time-travel.html): every way to read the past
- [Working with Transactions](transactions.html): commit metadata and several writers
- [Visualizing Your Lake](visualizing-your-lake.html): snapshot history, change volume, storage layout
- [Data Inlining](data-inlining.html): small writes and when to flush them
- [Choosing a Deployment](deployment.html): catalog backends, storage, and access for a team
- dplyneage's [column lineage for ducklake pipelines](https://tgerke.github.io/dplyneage/articles/ducklake-lineage.html), [project-level lineage for targets pipelines](https://tgerke.github.io/dplyneage/articles/targets-lineage.html), and [lineage that travels with the data](https://tgerke.github.io/dplyneage/articles/ducklake-versioned-lineage.html)

## References

- [CDISC SDTM](https://www.cdisc.org/standards/foundational/sdtm) and [CDISC ADaM](https://www.cdisc.org/standards/foundational/adam)
- [admiral](https://pharmaverse.github.io/admiral/) and [pharmaversesdtm](https://pharmaverse.github.io/pharmaversesdtm/)
- FDA, [Electronic Systems, Electronic Records, and Electronic Signatures in Clinical Investigations: Questions and Answers](https://www.fda.gov/regulatory-information/search-fda-guidance-documents/electronic-systems-electronic-records-and-electronic-signatures-clinical-investigations-questions) (October 2024)
- ICH, [E6(R3) Good Clinical Practice](https://www.ich.org/page/efficacy-guidelines)
- [DuckLake documentation](https://ducklake.select/docs/stable/)
